---
title: "Homework 3"
author: "Andrei Dmitrenko"
date: "March 12, 2019"
output:
    pdf_document:
html_document:
df_print: paged
---
  
# Task 6

## Posterior state probabilities

$P(Z_n = k\ |\ x) = \frac{P(Z_n=k,x)}{P(x)} = \frac{P(x_1,...,x_i,Z_n=k)P(x_{i+1},...,x_L\ |\ x_1,...,x_i,Z_n=k)}{P(x)} = \frac{P(x_1,...,x_i,Z_n=k)P(x_{i+1},...,x_L\ |\ Z_n=k)}{P(x)}$.  
The last equality holds due to the Markov property.

The proportionality constant is:  
$c = \frac{1}{P(x)} = \frac{1}{\sum_{Z_n} P(x,Z_n)}$

# Task 7

$P(x_{1:n}, Z_n=k) = P(x_{1:n-1},Z_n=k)\ P(x_n\ |\ x_{1:n-1}, Z_n=k)$


# Task 8

$P(x_{n+1:L}\ |\ Z_n = k) = \sum_j P(Z_{n+1} = j, x_{n+1:L}\ |\ Z_n = k)$.

Using chain rule:  
$P(Z_{n+1} = j, x_{n+1:L}\ |\ Z_n = k) = P(x_{n+2:L}, x_{n+1}, Z_{n+1}=k\ |\ Z_n=l)$  
$= P(x_{n+2:L}\ |\ x_{n+1}, Z_{n+1}=k, Z_n=l) \cdot P(x_{n+1}\ |\ Z_{n+1}=k, Z_n=l) \cdot P(Z_{n+1}=k\ |\ Z_n=l)$.

Simplifying:  
$P(Z_{n+1} = j, x_{n+1:L}\ |\ Z_n = k) = P(x_{n+2:L}\ |\ Z_{n+1}=k) \cdot P(x_{n+1}\ |\ Z_{n+1}=k) \cdot P(Z_{n+1}=k\ |\ Z_n=l)$.

Substituting defition of $\beta_{k}(n+1)$ we get:  
$\beta_k(n) = P(x_{n+1:L}\ |\ Z_n = k) = \sum_j P(Z_{n+1} = j\ |\ Z_n = k) \cdot P(x_{n+1}\ |\ Z_{n+1} = j) \cdot b_j(n+1),\ \Box$.


# Task 9

```{r}
#' Extract a specific emission probability from \data{emission}.
#'
#' @param state The state of the hidden variable.
#' @param observed The observed symbol.
#' @return The probability of `state` emitting `observed`.
#' @author Mathias Cardner
getEmission <- function(state, observed) {
  emission[state, observed]
}


#' Compute the forward quantity (message) at a given locus `n`.
#'
#' @param observed The observed symbol.
#' @param previous The forward quantity at locus `n-1`.
#' @return The forward quantity at the given locus.
#' @author Mathias Cardner
forward <- function(observed, previous) {
  # Compute the forward quantity using a matrix multiplication formulation.
  raw <- getEmission(c("+", "-"), observed) * transition %*% previous
  # Scale the result in order to prevent underflow.
  raw / sum(raw)
}

#' Compute the backward quantity (message) at a given locus `n`.
#'
#' @param observed The observed symbol.
#' @param upcoming The backward quantity at locus `n+1`.
#' @return The forward quantity at the given locus.
#' @author Mathias Cardner
backward <- function(observed, upcoming) {
  # Compute the backward quantity using a matrix multiplication formulation.
  raw <- transition %*% (getEmission(c("+", "-"), observed) * upcoming)
  # Scale the result in order to prevent underflow.
  raw / sum(raw)
}
```

## a) Forward quantities

```{r}
load("../data/HMM_GC_enrichment.RData")

# create a matrix for forward algorithm results
forward.values = matrix(rep(0, 2 * length(x)), nrow=2, ncol = length(x))
previous = c(0.5, 0.5)  # initial state

# compute other values with a forward algorithm
for (i in 1:length(x)){
  forward.values[,i] = forward(x[i], previous)
  previous = forward.values[,i]
}

forward.values[,1:10]
```

## b) Backward quantities

```{r}
# create a matrix for forward algorithm results
backward.values = matrix(rep(0, 2 * length(x)), nrow=2, ncol = length(x))
upcoming = c(1,1)  # initial state

# compute other values with a backward algorithm
for (j in length(x):1){
  backward.values[,j] = backward(x[j], upcoming)
  upcoming = backward.values[,j]
}

backward.values[,1:10]

```

## c) Posterior state

```{r}

# multiply results
multiplied.values = forward.values * backward.values
# scale results
scaled.multiplied.values = apply(multiplied.values, 2, function(x){ x / sum(x) })

scaled.multiplied.values[,1:10]

```

## d) Plot

```{r}
# plot
plot(Z == "+")
lines(Z == "+" & x == "C", col="lightblue")
```
