---
title: "Homework 5"
author: "Andrei Dmitrenko"
date: "March 28, 2019"
output:
    pdf_document:
html_document:
df_print: paged
---
  
# Task 11

## Connecting the transition matrix to the rate matrix

$\frac{d\textbf{P}(t)}{dt} = \frac{\textbf{P}(t+dt) - \textbf{P}(t)}{dt} = \frac{\textbf{P}(t)\cdot \textbf{P}(dt) - \textbf{P}(t)}{dt} =$  
$= \frac{\textbf{P}(t)\cdot (\textbf{P}(0)+\textbf{R}dt) - \textbf{P}(t)}{dt} = \frac{\textbf{P}(t)\cdot \textbf{P}(0) + \textbf{P}(t)\cdot \textbf{R}dt - \textbf{P}(t)}{dt} =$  
$= \textbf{P}(t)\cdot \textbf{R}+ \frac{\textbf{P}(t)\cdot \textbf{P}(0)}{dt} - \frac{\textbf{P}(t)}{dt} = \textbf{P}(t)\cdot \textbf{R}$, since $\textbf{P}(0)=\textbf{I}$.

# Task 12

## Connecting the rate matrix to the stationary distribution

$\textbf{R}\vec{\pi} = \vec{0}$.  
$\textbf{P}(dt)\vec{\pi} = (\textbf{P}(0)+\textbf{R}dt)\vec{\pi} = \vec{\pi} + \textbf{R}\vec{\pi}dt = \vec{\pi} = \pi$.  
The latter implies $\textbf{R}\vec{\pi} = \vec{0}$.

# Task 14

```{r}
library("phangorn")

data = read.dna("../data/ParisRT.txt")

# pairwise distances between sequences (Kimura model)
distance = dist.dna(data)

# create an initial tree topology for the alignment
initial.topology = nj(distance)

# plotting topology
plot(initial.topology)
```

```{r}
# fitting model to the data
fit = pml(initial.topology, data = phyDat(data), model = "K80")

# log-likelihood value
fit$logLik
```

```{r}
# optimise Q
fit.q = optim.pml(fit, optQ = TRUE)

# lower triangular part of the rate matrix
fit.q$Q

# the rate matrix
rm = matrix(0, 4, 4)
rm[upper.tri(rm, diag=FALSE)] <- fit.q$Q
rm[lower.tri(rm, diag=FALSE)] <- fit.q$Q
rm
```

```{r}
# optimise Q, edges, topology
fit.qet = optim.pml(fit, optQ = TRUE, optEdge = TRUE, optNni = TRUE)

# optimised log-likelihood
fit.qet$logLik
```

```{r results="hide"}
# bootstrap: resampling alignments to optimize edge weights and topology
bs <- bootstrap.pml(fit.qet, bs=100, optNni=TRUE)
```

```{r}
# plot shows: Mme S is more likely to have infected the patient Mme S 
treeBS <- plotBS(fit.qet$tree, bs, type = "phylogram")

```



