---
title: "Homework 4"
author: "Andrei Dmitrenko"
date: "March 18, 2019"
output:
    pdf_document:
html_document:
df_print: paged
---
  
# Task 8

## Estimating match emission probabilities

1. $e_1^M(\text{A}) = \frac{4+1}{(4+1)+(0+1)+(0+1)+(0+1)} = \frac{5}{8}$.  
$e_1^M(\text{C}) = e_1^M(\text{G}) = e_1^M(\text{T}) = \frac{0+1}{(4+1)+(0+1)+(0+1)+(0+1)} = \frac{1}{8}$.  

2. $e_2^M(\text{A}) = e_2^M(\text{C}) = e_2^M(\text{T}) = \frac{0+1}{(0+1)+(0+1)+(3+1)+(0+1)} = \frac{1}{7}$.  
$e_2^M(\text{G}) = \frac{3+1}{(0+1)+(0+1)+(3+1)+(0+1)} = \frac{4}{7}$.  

3. $e_3^M(\text{A}) = e_3^M(\text{G}) = e_3^M(\text{T}) = \frac{0+1}{(0+1)+(0+1)+(4+1)+(0+1)} = \frac{1}{8}$.  
$e_3^M(\text{C}) = \frac{4+1}{(0+1)+(0+1)+(4+1)+(0+1)} = \frac{5}{8}$.  


# Task 9

$e^I(\text{A}) = \frac{4+1}{(4+1)+(1+1)+(1+1)+(0+1)} =  \frac{1}{2}$.  
$e^I(\text{C}) = \frac{1+1}{(4+1)+(1+1)+(1+1)+(0+1)} =  \frac{1}{5}$.  
$e^I(\text{G}) = \frac{1+1}{(4+1)+(1+1)+(1+1)+(0+1)} =  \frac{1}{5}$.  
$e^I(\text{T}) = \frac{0+1}{(4+1)+(1+1)+(1+1)+(0+1)} =  \frac{1}{10}$.  


# Task 10

## Transitions for the sequences

- bat: $M_b \rightarrow M \rightarrow M \rightarrow I \rightarrow I \rightarrow I \rightarrow M \rightarrow M_e$.  
- rat: $M_b \rightarrow M \rightarrow D \rightarrow I \rightarrow I \rightarrow I \rightarrow M \rightarrow M_e$.  
- cat: $M_b \rightarrow M \rightarrow M \rightarrow I \rightarrow I \rightarrow I \rightarrow D \rightarrow M_e$.  
- gnat: $M_b \rightarrow D \rightarrow D \rightarrow I \rightarrow I \rightarrow I \rightarrow M \rightarrow M_e$.  
- gnat: $M_b \rightarrow M \rightarrow M \rightarrow I \rightarrow I \rightarrow I \rightarrow M \rightarrow M_e$.  

## State 0

Since the begin state (state 0) is assumed to be 'M', there are only 3 possible transitions:  

- $t_0(\text{M} \rightarrow \text{M}) = \frac{4+1}{(4+1)+(0+1)+(1+1)}= \frac{5}{8}$.   
- $t_0(\text{M} \rightarrow \text{I}) = \frac{0+1}{(4+1)+(0+1)+(1+1)}= \frac{1}{8}$.  
- $t_0(\text{M} \rightarrow \text{D}) = \frac{1+1}{(4+1)+(0+1)+(1+1)}= \frac{1}{4}$.  

## State 1

- $t_1(\text{M} \rightarrow \text{M}) = \frac{3+1}{(3+1)+(0+1)+(1+1)}= \frac{4}{7}$.  
- $t_1(\text{M} \rightarrow \text{I}) = \frac{0+1}{(3+1)+(0+1)+(1+1)}= \frac{1}{7}$.  
- $t_1(\text{M} \rightarrow \text{D}) = \frac{1+1}{(3+1)+(0+1)+(1+1)}= \frac{2}{7}$.  
- $t_1(\text{I} \rightarrow \text{M}) = \frac{0+1}{(0+1)+(0+1)+(0+1)}= \frac{1}{3}$.  
- $t_1(\text{I} \rightarrow \text{I}) = \frac{0+1}{(0+1)+(0+1)+(0+1)}= \frac{1}{3}$.  
- $t_1(\text{I} \rightarrow \text{D}) = \frac{0+1}{(0+1)+(0+1)+(0+1)}= \frac{1}{3}$.  
- $t_1(\text{D} \rightarrow \text{M}) = \frac{0+1}{(0+1)+(0+1)+(1+1)}= \frac{1}{4}$.  
- $t_1(\text{D} \rightarrow \text{I}) = \frac{0+1}{(0+1)+(0+1)+(1+1)}= \frac{1}{4}$.  
- $t_1(\text{D} \rightarrow \text{D}) = \frac{1+1}{(0+1)+(0+1)+(1+1)}= \frac{1}{2}$.  

## State 2

- $t_2(\text{M} \rightarrow \text{M}) = \frac{2+1}{(2+1)+(1+1)+(0+1)}= \frac{1}{2}$.  
- $t_2(\text{M} \rightarrow \text{I}) = \frac{1+1}{(2+1)+(1+1)+(0+1)}= \frac{1}{3}$.  
- $t_2(\text{M} \rightarrow \text{D}) = \frac{0+1}{(2+1)+(1+1)+(0+1)}= \frac{1}{6}$.  
- $t_2(\text{I} \rightarrow \text{M}) = \frac{2+1}{(2+1)+(3+1)+(1+1)}= \frac{1}{3}$.  
- $t_2(\text{I} \rightarrow \text{I}) = \frac{3+1}{(2+1)+(3+1)+(1+1)}= \frac{4}{9}$.  
- $t_2(\text{I} \rightarrow \text{D}) = \frac{1+1}{(2+1)+(3+1)+(1+1)}= \frac{2}{9}$.  
- $t_2(\text{D} \rightarrow \text{M}) = \frac{0+1}{(0+1)+(2+1)+(0+1)}= \frac{1}{5}$.  
- $t_2(\text{D} \rightarrow \text{I}) = \frac{2+1}{(0+1)+(2+1)+(0+1)}= \frac{3}{5}$.  
- $t_2(\text{D} \rightarrow \text{D}) = \frac{0+1}{(0+1)+(0+1)+(0+1)}= \frac{1}{5}$.  

## State 3

Since the end state (state 3) is assumed to be 'M', there are only 3 possible transitions:  

- $t_0(\text{M} \rightarrow \text{M}) = \frac{4+1}{(4+1)+(0+1)+(1+1)}= \frac{5}{8}$.   
- $t_0(\text{I} \rightarrow \text{M}) = \frac{0+1}{(4+1)+(0+1)+(1+1)}= \frac{1}{8}$.  
- $t_0(\text{D} \rightarrow \text{M}) = \frac{1+1}{(4+1)+(0+1)+(1+1)}= \frac{1}{4}$.  


# Task 11

```{r}
source("profileHMM.R")

# reading data
gtp.alignments = parseAlignment('../data/GTP_binding_proteins.txt')
atp.alignments = parseAlignment('../data/ATPases.txt')

# learning HMM parameters
gtp.hmm.params = learnHMM(gtp.alignments)
atp.hmm.params = learnHMM(atp.alignments)
```

```{r}
# plotting GTP HMM 
barplot(gtp.hmm.params$T[,50], main="GTP HMM transition probabilities", ylim = c(0,1))
barplot(gtp.hmm.params$mE[,50], main="GTP HMM emission probabilities", ylim = c(0,1))
barplot(gtp.hmm.params$iE[,50], main="GTP HMM insertion probabilities", ylim = c(0,1))
```

```{r}
# plotting ATP HMM 
barplot(atp.hmm.params$T[,50], main="ATP HMM transition probabilities", ylim = c(0,1))
barplot(atp.hmm.params$mE[,50], main="ATP HMM emission probabilities", ylim = c(0,1))
barplot(atp.hmm.params$iE[,50], main="ATP HMM insertion probabilities", ylim = c(0,1))
```

```{r}
# reading proteins data
proteins.data = parseProteins('../data/Unclassified_proteins.txt')

# applying forward algorithm to the data
gtp.hmm.log.ratios = c()
for (i in 1:length(proteins.data)){
  gtp.hmm.log.ratios = c(gtp.hmm.log.ratios, forward(gtp.hmm.params, proteins.data[[i]]))
}

# plotting log odds ratio for GTP HMM
plot(gtp.hmm.log.ratios, ylab = "Log odds ratio, GTP HMM", xaxt='n', xlab='Proteins')
lines(gtp.hmm.log.ratios)
axis(1, at = seq(1,31,1), las=2)
```

```{r}
# applying forward algorithm to the data
atp.hmm.log.ratios = c()
for (i in 1:length(proteins.data)){
  atp.hmm.log.ratios = c(atp.hmm.log.ratios, forward(atp.hmm.params, proteins.data[[i]]))
}

# plotting log odds ratio for GTP HMM
plot(atp.hmm.log.ratios, ylab = "Log odds ratio, ATP HMM", xaxt='n', xlab='Proteins')
lines(atp.hmm.log.ratios)
axis(1, at = seq(1,31,1), las=2)
```

```{r}
# plotting together
plot(atp.hmm.log.ratios, ylab = "Log odds ratio", xaxt='n', xlab='Proteins', ylim = c(-50,200), yaxt='n', col='blue', main='Log odds ratio of ATP and GTP HMMs separately')
lines(atp.hmm.log.ratios, col='blue')
points(gtp.hmm.log.ratios, ylab = "Log odds ratio", xaxt='n', xlab='Proteins', col='red')
lines(gtp.hmm.log.ratios, col='red')
axis(1, at = seq(1,31,1), las=2)
axis(2, at = seq(-50,200,50), las=2)
```

```{r}
# compute the difference
q_xs = gtp.hmm.log.ratios - atp.hmm.log.ratios

# plotting the q(x)
plot(q_xs, ylab = "q(x)", xaxt='n', xlab='Proteins', main="Log odds ratio for two models")
lines(q_xs)
axis(1, at = seq(1,31,1), las=2)
```

### Conclusions

- Proteins {1, 5, 9, 10, 12, 13, 14, 15, 16, 17, 22, 23, 25, 27, 28, 29} are likely to belong to ATPases family. The rest are likely to belong to GTP binding proteins.  

- Since the separation is quite clear, one can quite clearly decide for each protein.

