---
title: "Homework 9"
author: "Andrei Dmitrenko"
date: "April 24, 2019"
output:
    pdf_document:
html_document:
df_print: paged
---

# Task 30

## D-separation

1. $F,\ E$ are d-separated from A given ${G,\ D}$.  

2.    
A. true.  
B. false, since there's another path from G to E through C.  
C. false.  
D. true, since $MB(E) = F$.  

# Task 31

## Testing for marginal correlation

```{r}
data = readRDS("../data/MVN_DAG.rds")

# plotting shows no correlation
attach(data)
plot(A, B, main="Scatterplot", xlab="A", ylab="B")
```

We see no correlation. It agrees with the graph structure, since A | B.  

```{r}
# testing correlations
correlation.results = cor.test(data$A, data$B)

correlation.results$estimate
correlation.results$p.value
```

Test confirms there's no correlation.  

# Task 32

## Testing for partial correlation

```{r}
# regression A to C
ac.regression = lm(A ~ C, data = data)
ac.residuals = residuals(ac.regression)

# regression B to C
bc.regression = lm(B ~ C, data = data)
bc.residuals = residuals(bc.regression)

# plotting residuals
plot(ac.residuals, bc.residuals, main="Residuals", xlab="A ~ C", ylab="B ~ C")
```

I see weak negative correlation on the plot.  

```{r}
# correlation test for residuals
cor.test(ac.residuals, bc.residuals)
```

Test confirms, there's negative correlation of residuals, which agrees with the expectation,
since A and B are not independent given C.  

# Task 33

## Running the PC algorithm

```{r}
# PC algorithm
library(pcalg)

suffStat = list(C = cor(data), n = nrow(data))

pc.results.1 = pc(suffStat = suffStat, indepTest = gaussCItest, alpha = 1, labels = colnames(data), verbose=FALSE)
pc.results.2 = pc(suffStat = suffStat, indepTest = gaussCItest, alpha = 0.5, labels = colnames(data), verbose=FALSE)
pc.results.3 = pc(suffStat = suffStat, indepTest = gaussCItest, alpha = 0.2, labels = colnames(data), verbose=FALSE)
pc.results.4 = pc(suffStat = suffStat, indepTest = gaussCItest, alpha = 0.01, labels = colnames(data), verbose=TRUE)

par(mfrow=c(2,2))
plot(pc.results.1, main="Alpha = 1")
plot(pc.results.2, main="Alpha = 0.5")
plot(pc.results.3, main="Alpha = 0.25")
plot(pc.results.4, main="Alpha = 0.01")
```

The algorithm successfully learn the graph structure, provided a reasonable significance level alpha is specified. In this case the correct graph structure is obtained for $\alpha<0.15$. The structure obtained with $\alpha=0.01$ is shown on the plot.  

With increasing alpha we relax the threshold for correlation which results in highly interconnected graph structures.  






